Normalization by Evaluation for Sized Dependent Types

### 2017-02-27 nbe/icfp17.tex
Minor typos

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
%\thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

@@

In this article, we integrate dependent types and sized types with higher-rank size polymorphism, which is essential for generic programming and abstraction.  We introduce a size quantifier $\forall$ which lets us ignore sizes in terms for equality checking, alongside with a second quantifier $\Pi$ for abstracting over sizes that do affect the semantics of types and terms.  Judgmental equality is decided by an adaption of normalization-by-evaluation for our new type theory, which features \emph{type shape}-directed reflection and reification.  It follows that subtyping and type checking of normal forms are decidable as well, the latter by a bidirectional algorithm.
\end{abstract}

@@

% \cite{boveKraussSozeau:mscs16}
Consequently, each of the aforementioned languages based on Type Theory comes with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.  In current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.  This is a fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.  However, the syntactic termination check is very sensitive to reformulations of the program and hostile to abstraction \cite{abel:fics12}.

Sized types \cite{hughesParetoSabry:popl96} delegate the checking for structural descent to the type system by annotating data types with a size parameter.  The type checker can then ensure that in recursive calls, the size goes down, certifying termination.  In the simplest setting \cite{gimenez:typeBased,abel:lmcs07}, the size is just an upper bound on the tree height of the data structure; however, more sophisticated size annotations have also been considered \cite{xi:terminationHOSC,blanqui:rta04}.  Most sized type systems are non-dependent \cite{amadio:guardcondition,abelPientka:jfp16,bartheGregoireRiba:lernet08,bartheGregoireRiba:csl08,blanquiRiba:lpar06,dalLagoGrellois:probabilisticSizedTypes2017}, yet the combination of sized and dependent types has been studied as well \cite{blanqui:csl05,bartheGregoirePastawski:lpar06,gregoireSacchini:lpar10,sacchini:lics13,sacchini:flops14}.  However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types.\footnote{\citet{xi:terminationHOSC} has first-class size polymorphism, but only indexed types, no universes or large eliminations.}

Higher-rank size quantification takes termination checking to the next level; it is necessary for abstraction and generic programming.  For instance, it allows us to write a generic tree traversal which applies a user-given preprocessor on subtrees before recursively descending into these trees.  The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.  Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$ which takes a finite list of subtrees to form a new tree.
In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
In case $\vpre\,t = \tnode\,\vts$, the trees in the list $\vts$ are still guaranteed to be of strictly smaller size than $t$, thus, the recursive call to $\ttrav$, communicated via the $\tmap$ function for lists, is safe.
\[
\begin{array}{l}
  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i) (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
  \\
  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts)\, \})
\end{array}
\]
The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
\[
\begin{array}{l}
  \ttrav \; \vpre \; \vpost \; i\; t = \vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
\end{array}
\]
Church-style syntax is the basis for all program analyses and transformations to follow and show be considered as the \emph{true} syntax.
However, from a dependent-type perspective, explicit size applications
in terms can be problematic when the type checker compares terms for
equality---which is necessary as types can depend on values.  Inferred
sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for
$i \leq j$: we can always weaken an upper bound.  For instance, given
$\vts : \List\,(\tT\,i)$, any of the terms
$\tnode\,i\,\vts$, $\tnode\,(i+1)\,\vts$, $\dots$, $\tnode\,\infty\,\vts$
%$\tnode\,i\,\vts$ or $\tnode\,(i+1)\,\vts$ or $\dots$ or $\tnode\,\infty\,\vts$
has type
$\tT\,\infty$.  Yet semantically, all these trees are equal,
thus, the syntactic equality check should ignore the size argument to
$\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size
argument $i$ should be ignored by the equality check.  Yet
$\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have
different types for $i \not= j$, and moreover these function types are
not in the subtyping relation due to the mixed-variant occurrence of the
size parameter.  It seems that during equality checking we have to
consider terms of different types, at least for a while.  Once we apply
$\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which
determines $i = j = k$, we are back to safety.  However, allowing
types to differ during an equality check can be hazardous, especially
when the equality-check is type directed.

Consider the analogous situation for the polymorphic lambda calculus System~F, be it the predicative variant or not, extended by a unit type $\One$.
For Church-style, we can give a type-directed $\beta\eta$-equality test which equates all terms at the unit type.  The most interesting rules are the $\eta$-rules for unit and function type and the congruence rule for type application:
\begin{gather*}
\renewcommand{\jchk}{:}
\renewcommand{\jinf}{:}
  \ru{}{\Gamma \der t = t' \jchk \One}
\qquad
  \ru{\Gamma, x \of A \der t\,x = t'\,x \jchk B
    }{\Gamma \der t = t' \jchk A \to B}
\qquad
  \ru{\Gamma \der t = t' \jinf \forall X. B
    }{\Gamma \der t\,A = t'\,A \jinf B[A/X]}
\end{gather*}
The Curry-style version replaces the last conclusion by $\Gamma \der t = t' : B[A/X]$ where the type $A$ to instantiate $X$ has to be guessed.  However, in Curry-style more terms are equated than in Church-style, as for instance the Church-style terms $t\,A\,(\lambda x \of A.\, x)$ and $t\,B\,(\lambda x \of B.\, x)$ map to the same Curry-style term $t\,(\lambda x.\,x)$.  How would we adopt the algorithm for Church-style such that it equates all terms that are equal in Curry-style?  The conclusion of the last rule could be changed to $\Gamma \der t\,A = t'\,A' : B[A/X]$, but then the second term $t'\,A'$ does not have the ascribed type $B[A/X]$, and $\eta$-laws applied to this term would be unsound.  For instance, the algorithm would yield $t\,\One\,x = t\,(A \to A)\,y$ even for $x \not= y$.
We could also consider a heterogeneous check $(\Gamma \der t : A) = (\Gamma' \der t' : A')$ where each term is paired with its own type and context, but this leaves us with the dilemma of explaining the meaning of this judgement when $A$ and $A'$ are incompatible.

Does the literature offer a solution to this problem?
In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08,mishraLingerSheard:fossacs08} based on Miquel's Implicit Calculus of Constructions \citeyearpar{miquel:tlca01}.
In ICC$^*$, equality is checked by erasing all type abstractions and applications, and comparing the remaining untyped terms for $\beta\eta$-equality.  While this works for $\eta$-laws that can be formulated on untyped terms, such as $\eta$-contraction of functions $\lambda x.\,t\,x \redeta t$ (when $x$ not free in $t$), it does not extend to type-directed $\eta$-laws such as extensionality for the unit type.  Further, ICC$^*$ is not a type theory formulated with a typed equality judgement, which makes it hard to define its models \cite{miquel:lics00}---we wish not to go there, but stay within the framework of Martin-L\"of Type Theory \citeyearpar{martinlof:predicative75}.

%How can we have our cake and eat it, too?
Now, if the types of compared Curry-style terms are not equal, can they be sufficiently related to give a proper meaning to the algorithmic equality judgement?
It has already been observed that for a type-directed equality check the precise type is not necessary, a \emph{shape} or \emph{skeleton} is sufficient.  The skeleton informs the algorithm whether the terms under comparison are functions, inhabitants of the unit type, or something else, to possibly apply the appropriate $\eta$-law.  For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing the dependencies:  dependent function types map to non-dependent ones and indexed data types to simple data types. \citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to other type theories that admit dependency erasure.\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check. \citet{lovasPfenning:lmcs10} consider also \emph{refinement types for logical frameworks} which can be erased to simple types.}

By \emph{large eliminations} \cite{werner:types92} we refer to types computed by case distinction over values; they occur
in type theories that feature both universes and data types.
In the presence of large eliminations, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.  For instance consider the type $\funT n \NN \underbrace{A \to \dots \to A}_n \to A$ of $n$-ary functions; its shape is dependent on the value of $n$, thus cannot be determined statically.  Thus, the ``skeleton'' idea is also not directly applicable.

Going beyond the classical incremental equality check, there is a technique that can deal with dynamic $\eta$-expansion.  It is a type-directed normalization function inspired by normalization-by-evaluation (NbE) that computes $\eta$-long normal forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.  We can check the computed normal forms for identity and, thus, decide definitional equality.
NbE has proven to be a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.  It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.  At its heart there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.  Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.  Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.  For instance, the normal form of a closed function $f : U \to T$ would be $\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed variant $f : \funT x U T[x]$ it would be $\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.

% In the solution we present in this article, we do not introduce a new syntax for shapes, just a relation $T \shape S$ stating that type $S$ qualifies as shape for type $T$.  We use a normalization function inspired by normalization-by-evaluation (NbE) that computes the $\eta$-long normal form of the terms under comparison, which we then check for identity.

The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any \emph{shape} $S$ of $T$.
We managed, while not introducing a new syntax for shapes, to define a relation $T \shape S$ on type values stating that type $S$ qualifies as shape for type $T$.
Shapes unfold dynamically during reflection and reification.
For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
$(\up{\forall i.\/ F\,i}x)\,i = \up{F\,i} (x\,i)$ for size $i$ and
$(\up{\forall i.\/ F\,i}x)\,j = \up{F\,j} (x\,j)$ for size $j$.  The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.  This means they will still move in lock-step what $\eta$-expansion concerns, which is sufficient to prove NbE correct for judgmental equality.
We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.  Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.  This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.

For the time being, we do not (and cannot) develop a general theory of shape irrelevance.  We confine ourselves to size-irrelevant function types $\forall i.\, T[i]$.  This relieves us from defining a special shape-irrelevance modality, since all size-indexed types $T[i]$ \emph{are} shape irrelevant in $i$, simply because there is no case distinction on size, and sizes appear relevantly only under a sized type constructor such as $\tNat$.  Our technique would not extend to the polymorphic types $\forall X.\,B[X]$ of System~F.  Even though there is no case distinction on types, shape irrelevance of $B[X]$ fails in general, as $X$ could appear as a type on the top-level, e.g. in %$B[X] = X$ or
$B[X] = X \to X$, and then $B[\One]$ and $B[A \to A]$ would have distinct shapes.

To summarize, this article makes the following novel contributions:
\begin{enumerate}

@@

\item We adapt normalization-by-evaluation (NbE) to sized types and size quantification and show that it decides judgmental equality (sections~\ref{sec:sem} and \ref{sec:soundnbe}).  The novel technical tool is a relation $T \shape S$ which relates a type $T$ to its possible shapes $S$.  This approximation relation allows reflection and reification at size-polymorphic types $\forall i.\, T$.
As usual for the meta-theory of Type Theory with large eliminations, the machinery is involved, but we just require the usual two logical relations:  First, a PER model to define the semantics of types and prove the completeness of NbE (Section~\ref{sec:sem}).  Secondly, a relation between syntax and semantics to prove soundness of NbE (Section~\ref{sec:soundnbe}).

\item We present an bidirectional type checking algorithm \cite{coquand:type} which takes the irrelevant size argument as reliable hint for the type checker (sections~\ref{sec:asub} and \ref{sec:tycheck}).  It is complete for normal forms which can be typed with the restricted rule for $\forall$-elimination:
\[
\ru{\Gamma \der t : \forall i.\, T
  }{\Gamma \der t\,a : T[a/i]}
\]
The algorithm employs the usual lazy reduction for types, \ie, just-in-time weak-head evaluation, in type and sub-type checker
\cite{huet:wsscs89}.  In this, it improves on \citet{fridlenderPagano:tlca13} which instruments full normalization (NbE) at every step.
\end{enumerate}
This article is accompanied by a prototypical type checker \textsf{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
But before going into the technical details, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.

@@

\section{A Type System With Irrelevant Size Application}
\label{sec:syntax}

@@

We extend the syntax by de Bruijn levels $\lev k$
to be used as generic values (unknowns) and type annotations
$\up T t$ and $\down T t$ for lazy realizations of the
reflection and reification operations of NbE.
\[
\begin{array}{lllrl@{\qquad}l}
%  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
  \Ne   & \ni & n & ::= & \dots \mid \lev k \\
  \tUp  & \ni & N & ::= & \up T n \\
  \Whnf & \ni & w & ::= & \dots \mid N \\
  \tDown & \ni & d & ::= & \down T t \\
  \Exp  & \ni & t & ::= & \dots \mid d \\
%  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
\end{array}
\]

@@

Figure~\ref{fig:typedchart}, adapted from \citet{abel:habil} shows the syntactic categories and main operations involved in NbE in what we call \emph{locally nameless style}.  The red path $\Exp \to \D \to \Dnf \to \Nf$ decomposes $\beta\eta$-normalization into three steps.
\begin{enumerate}
\item
First, we close the term $t$ with an environment $\eta$ that maps the free de Bruijn indices of $t$ to reflected de Bruijn levels.  Reflection of de Bruijn levels follows the blue path $\Level \to \Dne \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.

@@

Note that $T \in \SET[\ell]$ and $T \shape S$ do not imply $S \in \SET[\ell]$.
Type shapes are not well-defined types in general.
For instance, assume a term $F : \Nat 0 \to \Set[0]$ which diverges if applied to a successor term.
Then $T := \funT x {\Nat 0} {F\,x}$ is a well-defined type; we have $T \in \SET[0]$.
Now consider $S := \funT x {\Nat \infty} {F\,x}$.
We have $T \shape S$, but not $S$ is not well-defined; $S \not\in \SET[0]$.
% $\funT x {\Nat 0} {F\,x} \shape , but the latter one is not a well-defined type.

>>>>

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
\thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

@@

In this article, we integrate dependent types and sized types with higher-rank size polymorphism, which is essential for generic programming and abstraction.  We introduce a size quantifier $\forall$ which lets us ignore sizes in terms for equality checking, alongside with a second quantifier $\Pi$ for abstracting over sizes that do affect the semantics of types and terms.  Judgemental equality is decided by an adaption of normalization-by-evaluation for our new type theory, which features \emph{type shape}-directed reflection and reification.  It follows that subtyping and type checking of normal forms are decidable as well, the latter by a bidirectional algorithm.
\end{abstract}

@@

% \cite{boveKraussSozeau:mscs16}
Consequently, each of the aforementioned languages based on Type Theory come with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.  In the current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.  This fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.  However, the syntactic termination check is very sensitive to reformulations of a program and hostile to abstraction \cite{abel:fics12}.

Sized types \cite{hughesParetoSabry:popl96} delegate the checking for structural descent to the type system by annotating data types with a size parameter.  The type checker can then ensure that in recursive calls, the size goes down, certifying termination.  In the simplest setting \cite{gimenez:typeBased,abel:lmcs07}, the size is just an upper bound on the tree height of the data structure; however, more sophisticated size annotations have also been considered \cite{xi:terminationHOSC,blanqui:rta04}.  Most sized type systems are non-dependent \cite{amadio:guardcondition,abelPientka:jfp16,bartheGregoireRiba:lernet08,bartheGregoireRiba:csl08,blanquiRiba:lpar06,dalLagoGrellois:probabilisticSizedTypes2017}, yet the combination of sized and dependent types has been studied as well \cite{blanqui:csl05,bartheGregoirePastawski:lpar06,gregoireSacchini:lpar10,sacchini:lics13,sacchini:flops14}.  However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types (Xi \citeyearpar{xi:terminationHOSC} only has indexed types).

Higher-rank size quantification takes termination checking to the next level; it is necessary for abstraction and generic programming.  For instance, it allows us to write a generic tree traversal which applies a user-given preprocessor on subtrees before recursively descending into these trees.  The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.  Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$.
In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
In case $\vpre\,t = \tnode\,\vts$, the trees in the list $\vts$ are still guaranteed to be of strictly smaller size than $t$, thus, the recursive call to $\ttrav$, communicated via the $\tmap$ function for lists, is safe.
\[
\begin{array}{l}
  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i) (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
  \\
  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts) \})
\end{array}
\]
This is the \emph{Curry}-style program as accepted from the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
\[
\begin{array}{l}
  \ttrav \; \vpre \; \vpost \; i\; t = \vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
\end{array}
\]
From a dependent-type perspective, explicit size applications in terms can be problematic when the type checker compares terms for equality (which is necessary as types can depend on values).  Inferred sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for $i \leq j$: we can always weaken an upper bound.  For instance, given $\vts : \List\,(\tT\,i)$, all of $\tnode\,i\,\vts$ and $\tnode\,(i+1)\,\vts$ and $\dots$ and $\tnode\,\infty\,\vts$ have type $\tT\,\infty$.  However, semantically, all these trees are equal, thus, the syntactic equality check should ignore the size argument to $\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size argument $i$ should be ignored by the equality check.  Yet $\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have different types for $i \not= j$, and moreover these function types are not in the subtyping relation due to mixed-variant occurrence of the size parameter.  It seems that during equality checking we have to consider terms of different types, for a while.  Once we apply $\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which determines $i = j = k$, we are back in safety.  However, allowing types to differ during an equality check can be hazardous, especially when the equality-check is type directed.

Consider the analogous situation for the polymorphic lambda calculus System~F, be it the predicative variant or not, extended by a unit type $\One$.
For Church-style, we can give a type-directed $\beta\eta$-equality test which equates all terms at the unit type.  The most interesting rules would be the $\eta$-rules for unit and function type and the congruence rule for type application:
\begin{gather*}
\renewcommand{\jchk}{:}
\renewcommand{\jinf}{:}
  \ru{}{\Gamma \der t = t' \jchk \One}
\qquad
  \ru{\Gamma, x \of A \der t\,x = t'\,x \jchk B
    }{\Gamma \der t = t' \jchk A \to B}
\qquad
  \ru{\Gamma \der t = t' \jinf \forall X. B
    }{\Gamma \der t\,A = t'\,A \jinf B[A/X]}
\end{gather*}
The Curry-style version would replace the last conclusion by $\Gamma \der t = t' : B[A/X]$ where the type $A$ to instantiate $X$ has to be guessed.  However, in Curry-style more terms are equated than in Church-style, as for instance the Church-style terms $t\,A\,(\lambda x \of A.\, x)$ and $t\,B\,(\lambda x \of B.\, x)$ map to the same Curry-style term $t\,(\lambda x.\,x)$.  How would we adopt the algorithm for Church-style such that it equates all terms that are equal in Curry-style?  The conclusion of the last rule could be changed to $\Gamma \der t\,A = t'\,A' : B[A/X]$, but then the second term $t'\,A'$ does not have the ascribed type $B[A/X]$, and $\eta$-laws applied to this term would be unsound.  For instance, the algorithm would yield $t\,\One\,x = t\,(A \to A)\,y$ even for $x \not= y$.
We could also consider a heterogeneous check $(\Gamma \der t : A) = (\Gamma' \der t' : A')$ where each term is paired with its own type and context, but this leaves us with the dilemma of explaining the meaning of this judgement when $A$ and $A'$ are incompatible.

Does the literature offer a solution to this problem?
In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08,mishraLingerSheard:fossacs08} based on Miquel's Implicit Calculus of Constructions \cite{miquel:tlca01}.
In ICC$^*$, equality is checked by erasing all type abstractions and applications, and comparing the remaining untyped terms for $\beta\eta$-equality.  While this works for $\eta$-laws that can be formulated on untyped terms, such as $\eta$-contraction of functions $\lambda x.\,t\,x \redeta t$ (when $x$ not free in $t$), it does not extend to type-directed $\eta$-laws such as extensionality for the unit type.  Further, ICC$^*$ is not a type theory formulated with a typed equality judgement, which makes it hard to define its models \cite{miquel:lics00}---we wish not to go there, but stay within the framework of Martin-L\"of Type Theory \cite{martinlof:predicative75}.

How can we have our cake and eat it, too?
It has already been observed that for a type-directed equality check the precise type is not necessary, a \emph{shape} or \emph{skeleton} is sufficient that informs the algorithm whether the terms under comparison are functions, inhabitants of the unit type, or something else.  For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing all dependencies. \citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to all type theories that admit dependency erasure.\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check. \cite{lovasPfenning:lmcs10} consider also \emph{refinement types for logical frameworks} which can be erased to simple types.}
In the presence of large eliminations, which we have in type theories that feature both universes and data types, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.  For instance consider the type $\funT n \NN \underbrace{A \to \dots \to A}_n \to A$ of $n$-ary functions; its shape is dependent on the value of $n$, thus cannot be determined statically.  In the solution we present in this article, we do not introduce a new syntax for shapes, just a relation $T \shape S$ stating that type $S$ qualifies as shape for type $T$.  We use a normalization function inspired by normalization-by-evaluation (NbE) that computes the $\eta$-long normal form of the terms under comparison, which we then check for identity.  NbE \cite{bergerSchwichtenberg:lics91,danvy:tdpe} has proven a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.  It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.  At its hard there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.  Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and lets us compute with open terms.  Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.  For instance, the normal form of a closed function $f : U \to T$ would be $\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed variant $f : \funT x U T[x]$ it would be $\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.

The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any shape $S$ of $T$.  For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
$(\up{\forall i.\/ F\,i}x)\,i = \up{F\,i} (x\,i)$ for size $i$ and
$(\up{\forall i.\/ F\,i}x)\,j = \up{F\,j} (x\,j)$ for size $j$.  The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.  This means they will still move in lock-step what $\eta$-expansion concerns, which is sufficient to prove NbE correct for judgmental equality.
We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.  Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.  This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is (at least, currently) out of reach, and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.

For the time being, we do not (and cannot) develop a general theory of shape irrelevance.  We confine ourselves to size-irrelevant function types $\forall i.\, T[i]$.  This relieves us from defining a special shape-irrelevance modality, since all size-indexed types $T[i]$ are shape irrelevant in $i$, simply because there is no case distinction on size, and sizes appear relevantly only under a sized type constructor such as $\tNat$.  Our technique would not extend to the polymorphic types $\forall X.\,B[X]$ of System~F.  Even though there is no case distinction on types, shape irrelevance of $B[X]$ fails in general, as $X$ could appear as a type on the top-level, e.g. in $B[X] = X$ or $B[X] = X \to X$.

This article makes the following novel contributions:
\begin{enumerate}

@@

\item We adapt normalization-by-evaluation (NbE) to sized types and size quantification and show that it decides judgmental equality (sections~\ref{sec:sem} and \ref{sec:soundnbe}).  The novel technical tool is a relation $T \shape S$ which relates a type $T$ to its possible shapes $S$.  This approximation relation allows reflection and reification at size-polymorphic types $\forall i.\, T$.
As usual for the meta-theory of Type Theory with large eliminations, the machinery is involved, but just require the usual two logical relations:  First, a PER model to define the semantics of types and prove the completeness of NbE (Section~\ref{sec:sem}).  Secondly, a relation between syntax and semantics to prove soundness of NbE (Section~\ref{sec:soundnbe}).

\item We present an bidirectional type checking algorithm \cite{coquand:type} which takes the irrelevant size argument as reliable hint for the type checker (sections~\ref{sec:asub} and \ref{sec:tycheck}).  It is complete for normal forms which can be typed with the restricted rule for $\forall$-elimination:
\[
\ru{\Gamma \der t : \forall i.\, T
  }{\Gamma \der t\,a : T[a/i]}
\]
The algorithm employs the usual lazy reduction for types (just-in-time weak-head evaluation) in type and sub-type checker
\cite{huet:wsscs89}.  In this, it improves on \citet{fridlenderPagano:tlca13} which instruments full normalization (NbE) at every step.
\end{enumerate}
This article is accompanied by a prototypical type checker \texttt{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
But first, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.

@@

\section{A Type System WIth Irrelevant Size Application}
\label{sec:syntax}

@@

We extend the syntax by de Bruijn levels $\lev k$
to be used as generic values (unknowns) and type annotations
$\up T t$ and $\down T t$ for lazy realisations of the
reflection and reification operations of NbE.
\[
\begin{array}{lllrl@{\qquad}l}
%  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
  \Ne   & \ni & n & ::= & \dots \mid \lev k \\
  \tUp  & \ni & N & ::= & \up T n \\
  \Whnf & \ni & w & ::= & \dots \mid N \\
  \tDown & \ni & d & ::= & \down T t \\
  \Exp  & \ni & t & ::= & \dots \mid d \\
%  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
\end{array}
\]

@@

Figure~\ref{fig:typedchart}, adapted from \citet{abel:habil} shows the syntactic categories and main operations involved in NbE in what we call \emph{locally nameless style}.  The red path $\Exp \to \D \to \Dnf \to \Nf$ decomposes $\beta\eta$-normalization into three steps.
\begin{enumerate}
\item
First, we close the term $t$ with an environment $\eta$ that maps the free de Bruijn indices of $t$ to reflected de Bruijn levels.  Reflection of de Bruijn levels follows the blue path $\Level \to \Dne \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labelled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.

@@

Note that $T \in \SET[\ell]$ and $T \shape S$ do not imply $S \in \SET[\ell]$.
Type shapes are not well-defined types in general.
For instance, assume a term $F : \Nat 0 \to \Set[0]$ which diverges if applied to a successor term.
Then $T := \funT x {\Nat 0} {F\,x}$ is a welldefined type; we have $T \in \SET[0]$.
Now consider $S := \funT x {\Nat \infty} {F\,x}$.
We have $T \shape S$, but not $S$ is not well-defined; $S \not\in \SET[0]$.
% $\funT x {\Nat 0} {F\,x} \shape , but the latter one is not a well-defined type.
